From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-04-12 appendixB/appendixB.Rmd
WIP

>>>>

## Summary of more complex models

There

Reference        |  Elements                   |  Description
---------------- | ----------------------------|-----------------
@Cohen1979       | Age/stage structure
@Tuljapurkar1989 | Age/stage structure 
@Tuljapurkar1997 | Age/stage structure 
@Caswell2009     | Age/stage structure
@Metcalf2015     | Age/stage structure 
@Vindenes2017b   | Age/stage structure
@Durrett1994     | Spatial structure
@Dieckmann2000   | Spatial structure
@Kerr2002        | Spatial structure
@Schreiber2010   | Spatial structure
@Vindenes2015    | Individual heterogeneity 
@Hart2016        | Individual heterogeneity 
@Caswell1978     | Coexistence
@Tuljapurkar1980 | Coexistence
@Chesson1981     | Coexistence
@Chesson1982     | Coexistence
@Chesson1985     | Coexistence
@Chesson1989     | Coexistence
@Melbourne2007   | Coexistence
@Schreiber2017   | Coexistence
@Ripa            | Colored noise
@Roughgarden     | Colored noise
@Ripa1999        | Colored noise
@Petchey         | Colored noise
@Schreiber2010   | Colored noise
@Lee2017         | Colored noise 
@Spanio2017      | Colored noise 
@Coulson2001     | periodic/temporal structure
@Bjornstad2001   | periodic/temporal structure
@Keeling2001     | periodic/temporal structure
@Saether1997     | eco-evolutionary
@Ozgul2009       | eco-evolutionary
@Lenormand2009   | eco-evolutionary
@Vindenes2015    | eco-evolutionary
@Schreiber2015   | eco-evolutionary

### 2018-04-12 paper/paper.Rmd
WIP

To the empirical ecologist, stochasticity, or more simply, noise,
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of statistical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuisance
that obscures the deterministic skeleton of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transition from *noise the nuisance*
to *noise the creator* of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empirical thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become *noise the informer*.  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empirical communities by 
seeing noise not as mathematical curiosity or statistical bugbear, but as a
source for new opportunities for inference.

The gap between specialists and non-specialists is particularly acute in the
ecological literature on stochastic processes.  This makes the potential benefit
for new synthesis all the greater, but also poses an obvious challenge in 
appealing to both audiences.  In recognition of this, I provide concise and
commented code for simulating each of the models presented in the supplement,
Appendix A, while more mathematical background and derivations are presented Appendix B.
Numerical simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.

@@

Empirical evidence for stochastic switching in ecological timeseries
is limited by the long time scale usually required 
[but see @Keeling2001; @Taylor1993; @Wang2012; @Clements2016; @Clements2017]. 
As our discussion highlights, we are most likely to observe the phenomenon when at
least one of the potential wells is relatively shallow, which
results in a pattern of long periods in a single state (the deeper well)
punctuated by brief excursions into and back out of the shallow well.
This drives the more spike-like pattern seen in the rubella outbreaks
from @Keeling2001. 
Stochastic transitions between bi-stable states are thought to play
an important role in genetic switches, where among other phenomena
this mechanism can explain the emergence of two alternative phenotypes
in clonal microbial populations, which can act to buffer the
population in a changing environment  [@Balaban2004]. This mechanism
has also been explored in the emergence of multiple strategies
under evolutionary dynamics in disease models [@Read2007].

>>>>

To many, stochasticity, or more simply, noise,
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of statistical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, literature in stochasticity has transitioned
from thinking of stochasticity in such terms; where noise is a nuisance
that obscures the deterministic skeleton of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transition from *noise the nuisance*
to *noise the creator* of ecological phenomena has had, with a few notable exceptions,
relatively little impact in broader thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become *noise the informer*.  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empirical communities by 
seeing noise not as mathematical curiosity or statistical bugbear, but as a
source for new opportunities for inference.

This distinction between a paradigms of noise the "nuisance" versus noise the "creator"
is necessarily a simplification across a spectrum of perspectives. These categories
should not be mistaken for either a sharp dichotomy nor a reference to a stictly
empirical-theoretical divide.  It is also essential to recognize that each paradigm
expands upon rather than rejects the previous notion of noise: the recognition that
noise can create novel phenomena does not mean that noise cannot also obscure the 
signal of some process of interest.  Likewise, seeking to use noise as a novel source
of information about underlying processes will be informed by both previous paradigms,
as our discussion will illustrate.

Accompanying this discussion, I provide concise and
commented code for simulating each of the models we will discuss as
Appendix A, while more mathematical background and derivations can be found in Appendix B.
Numerical simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.

@@

Empirical evidence for stochastic switching in ecological timeseries
is limited by the long time scale usually required 
[but see @Keeling2001; @Taylor1993; @Wang2012; @Clements2016; @Clements2017]. 
As our discussion highlights, we are most likely to observe the phenomenon when at
least one of the potential wells is relatively shallow, which
results in a pattern of long periods in a single state (the deeper well)
punctuated by brief excursions into and back out of the shallow well.
This drives the more spike-like pattern seen in the rubella outbreaks
from @Keeling2001. 
Stochastic transitions between bi-stable states are thought to play
an important role in genetic switches, where among other phenomena
this mechanism can explain the emergence of two alternative phenotypes
in clonal microbial populations, which can act to buffer the
population in a changing environment  [@Balaban2004]. This mechanism
has also been explored in the emergence of multiple strategies
under evolutionary dynamics in disease models [@Read2007].
This further underscores that the collection of more long-term data
will essential to our understanding noisy phenomena and their underlying mechanisms.
